tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-14 paper.tex
Update on performance

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Fig~\ref{fig:time}A shows how the simulation time scales
with the number of individuals using simple simulations
of humans using \texttt{stdpopsim}~\citep{adrion2020}.
[HERE]
The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
VCF, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.

>>>>

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Fig~\ref{fig:time}A shows how the trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
For comparison, we also report the CPU time required by a
modified version of the code used by \citet{martin2017} for a
subset of these simulated ARGs.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the time required to read the data at this scale would be
far more than the minutes required here to perform
the simulations.
We do not report the memory usage here, but note that these
simulations for up to 5 million humans were run without difficult
on a laptop.

For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)

[HERE]
The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
VCF, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.
